---
title: "collaborative-data-science"
format: html
editor: visual
---

# Introduction

High-dimensional proteomics data generated from mass spectrometry often includes thousands of proteins measured across multiple experimental conditions. In studies involving different treatments and intoxication states, it can be difficult to directly interpret patterns in protein abundance due to the complexity of the dataset.

Principal component analysis (PCA) is a commonly used method to reduce the dimensionality of this type of data while preserving the most important sources of variation. By transforming the original variables into a smaller set of principal components, PCA allows researchers to more easily visualize and interpret differences between experimental conditions.

# Objective

The objective of this package is to provide a simple and flexible workflow for analyzing high-dimensional proteomics data. Specifically, this package allows users to:

-   Perform principal component analysis (PCA) on numeric datasets\
-   Determine the optimal number of principal components using scree plots\
-   Visualize variance explained across components\
-   Perform cluster analysis to group similar observations

These tools are designed to help identify patterns in protein abundance that may be associated with treatment or intoxication conditions.

# Workflow Overview

The general workflow for this package includes:

1.  Input a dataset containing protein abundance values\
2.  Perform PCA using the `pca()` function\
3.  Generate scree plots using `scree_plot()` to determine the number of meaningful components\
4.  Apply clustering using `cluster_analysis()` to identify groups within the data

This workflow helps simplify complex datasets and highlights underlying structure in the data.

```{r}
#library(dplyr)
library(ggplot2)
#iris <- read.csv("data/Iris.csv") %>%
 # select(-Id) # example dataset
#detach("package:dplyr", unload = TRUE)
```

```{r}
scree_plot <- function(eigen_values, threshold = 85, y_axis = "Eigenvalue") { # generates a scree plot and determines how many principal components are optimal for analysis
  num_comp <- 0
  exp_var <- 0
  while(exp_var < threshold) { # determines the minimum number of principal components that explains the minimal threshold percent of variance (default is 85 percent)
    num_comp <- num_comp + 1
    exp_var <- exp_var + 100 * eigen_values[num_comp]/sum(eigen_values)
  }
  percentage <- t(as.matrix(100 * eigen_values/sum(eigen_values))) # determines how much of the dataset's variance can be explained by the given eigenvalues
  columns <- c()
  for(i in 1:ncol(percentage)) { # name columns to their appropriate principal components
    columns <- append(columns, paste0("PC", i))
  }
  colnames(percentage) <- columns
  if(y_axis == "Percentage") { # y axis is the actual eigenvalue of each number of components, but can be modified to show percentage of explained variance instead
    y_axis <- "Percentage of Variance Explained"
    eigen_values <- percentage[1,]
  }
  value_coords <- data.frame(1:length(eigen_values), eigen_values)
  plot <- ggplot2::ggplot(data = value_coords, aes(x = value_coords[, 1], y = value_coords[, 2])) +
    ggplot2::geom_point() +
    ggplot2::geom_line() +
    ggplot2::xlab("Principal Component") +
    ggplot2::ylab(y_axis) +
    ggplot2::geom_vline(xintercept = num_comp, col = "red") +
    ggplot2::theme(panel.border = element_rect(colour = "black"))
  output <- list(plot, num_comp, percentage)
  names(output) <- c("plot", "optimal_number_of_components", "percentage_of_variance_explained")
  return(output)
}
```

```{r}
pca <- function(data, eigen_opt = NA, threshold = 85) { # conducts principal component analysis
  std <- data[0, sapply(data, is.numeric)] # exclude non-numeric properties from analysis
  for(i in colnames(std)) { # standardized values for each factor
    std[1, i] <- mean(as.numeric(data[, i]))
    std[2, i] <- sd(as.numeric(data[, i]))
  }
  rownames(std) <- c("mean", "sd")
  trns <- data[, sapply(data, is.numeric)] # transform (AKA standardize) data so that each factor has a mean of 0 and a variance of 1
  for(i in colnames(std)){
    trns[, i] <- (data[, i] - std[1, i])/std[2, i]
  }
  cov <- data[0, sapply(data, is.numeric)] # generate a covariance matrix for each factor
  for(i in colnames(cov)) { # i is column and j is row, so should go [j, i] 
    for(j in colnames(cov)) {
      cov[j, i] <- cov(trns[, j], trns[, i])
    }
  }
  eigen <- eigen(data.matrix(cov)) # generate eigenvectors and eigenvalues from the covariance matrix
  if(is.na(eigen_opt)) { # if the desired number of principal components is not specified, this will be determined by the "optimal" number of principal components based on "threshold" (that is, what is the minimum number of principal components needed to explain "threshold" % of the dataset ("threshold" default is 85%))
    eigen_opt <- scree_plot(eigen$values, threshold = threshold)$optimal_number_of_components
  }
  pca <- as.matrix(trns[, sapply(trns, is.numeric)]) %*% eigen$vectors[, 1:eigen_opt] # conduct a matrix mutiplication to transform the original data into principal components
  columns <- c()
  for(i in 1:ncol(pca)) { # name columns to their appropriate principal components
    columns <- append(columns, paste0("PC", i))
  }
  colnames(pca) <- columns
  output <- list(pca, eigen_opt, eigen$values, eigen$vectors, as.matrix(cov), as.matrix(trns), as.matrix(std))
  names(output) <- c("pca", "optimal_number_of_principal_components", "eigen_values", "eigen_vectors", "covariance_matrix", "transformed_data", "standardized_values")
  return(output)
}
```

```{r}
test <- pca(iris)
head(test$pca)
```

```{r}
test$optimal_number_of_principal_components
```

```{r}
test$eigen_values
```

```{r}
test$eigen_vectors
```

```{r}
test$covariance_matrix
```

```{r}
head(test$transformed_data)
```

```{r}
test$standardized_values
```

```{r}
testplot <- scree_plot(test$eigen_values)
testplot$plot
```

```{r}
testplot <- scree_plot(test$eigen_values, y_axis = "Percentage")
testplot$plot
```

```{r}
testplot$optimal_number_of_components
```

```{r}
testplot$percentage_of_variance_explained
```

```{r}
cluster_analysis <- function(pca, num_clusters = NA, max_clusters = 10, x = "PC1", y = "PC2", nstart = 10) { # segregates points from the principal component analysis into different clusters using kmeans clustering
  twss <- c()
  for(i in 1:max_clusters) { # generates values for the total within-cluster sum of squares; important for this method of determining "optimal" number of clusters
    twss <- append(twss, kmeans(pca$pca, centers = i, nstart = nstart)$tot.withinss)
  }
  clust_coords <- data.frame(1:max_clusters, twss)
  clust_opt <- 1
  while(isTRUE(twss[clust_opt] > twss[clust_opt + 1])) { # for this method, the "optimal" number of clusters is determined by the "elbow" of the scree plot--in other words, what is the smallest number of clusters whose total within-cluster sum of squares is a local minimum?
    clust_opt <- clust_opt + 1
  }
  if(is.na(num_clusters)) { # if no desired number of clusters is given, it defaults to the "optimal" number of clusters
    num_clusters <- clust_opt
  }
  kmeans <- kmeans(pca$pca, centers = num_clusters, nstart = nstart)
  cluster_values <- kmeans$cluster
  qual <- 100 * kmeans$betweenss / kmeans$totss
  cluster_determine <- ggplot2::ggplot(data = clust_coords, aes(x = clust_coords[, 1], y = clust_coords[, 2])) +
    ggplot2::geom_point() +
    ggplot2::geom_line() +
    ggplot2::xlab("Number of Clusters") +
    ggplot2::ylab("Total Within Sum of Squares") +
    ggplot2::geom_vline(xintercept = clust_opt, col = "red") +
    ggplot2::theme(panel.border = element_rect(colour = "black"))
  cluster_scatter <- ggplot2::ggplot(data = pca$pca, aes(x = pca$pca[, x], y = pca$pca[, y], color = cluster_values)) +
    ggplot2::geom_point() +
    ggplot2::xlab(x) +
    ggplot2::ylab(y) +
    ggplot2::theme(panel.border = element_rect(colour = "black"), aspect.ratio = 1)
  output <- list(cluster_values, clust_opt, qual, cluster_determine, cluster_scatter)
  names(output) <- c("cluster_values", "optimal_number_of_clusters", "quality", "optimal_cluster_plot", "pca_scatterplot")
  return(output)
}
```

```{r}
testclust <- cluster_analysis(test)
testclust$cluster_values
```

```{r}
testclust$optimal_number_of_clusters
```

```{r}
testclust$quality
```

```{r}
testclust$optimal_cluster_plot
```

```{r}
testclust$pca_scatterplot
```

```{r}
testclust2 <- cluster_analysis(test, num_clusters = 2, nstart = 1)
testclust2$cluster_values
```

```{r}
testclust2$optimal_number_of_clusters
```

```{r}
testclust$quality
```

```{r}
testclust2$optimal_cluster_plot
```

```{r}
testclust2$pca_scatterplot
```

# Conclusion

This package provides a straightforward approach for analyzing high-dimensional proteomics data using principal component analysis and clustering methods. By reducing dimensionality and identifying patterns in protein abundance, these tools help simplify complex datasets and highlight meaningful differences across experimental conditions such as treatment and intoxication.

Overall, this workflow supports more efficient exploration and interpretation of large-scale proteomics data.
